---
title: "Individual Sample Analysis"
author: "Jean Fan"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = 'figure/',
    cache.path = 'cache/',
    cache = TRUE,
    dev = 'png',
    dpi=100, fig.width=6, fig.height=3
)
```

To demonstrate the utility of `MUDAN` for analysis of an individual sample, we will use simulated data to show how poor signal can be strengthened by graph-based community detection. 

Let's begin by simulating some data where we have 5 different cell types / groups.

```{r}
library(MUDAN)

## simulate data
simulate.data <- function(G=5, N=100, M=1000, initmean=0, initvar=10, upreg=1, upregvar=15, ng=10, ng2=20, seed=0, plot=TRUE) {
  set.seed(seed)
  mat <- matrix(rnorm(N*M*G, initmean, initvar), M, N*G)
  rownames(mat) <- paste0('gene', 1:M)
  colnames(mat) <- paste0('cell', 1:(N*G))
  group <- factor(sapply(1:G, function(x) rep(paste0('group', x), N)))
  names(group) <- colnames(mat)
  
  ## unique diff genes
  diff <- lapply(1:G, function(x) {
    diff <- rownames(mat)[(((x-1)*ng)+1):(((x-1)*ng)+ng)]
    mat[diff, group==paste0('group', x)] <<- mat[diff, group==paste0('group', x)] + rnorm(ng, upreg, upregvar)
    return(diff)
  })
  names(diff) <- paste0('group', 1:G)
  
  ## shared subpops
  diff2 <- lapply(2:(G-1), function(x) {
    y <- x+G
    diff <- rownames(mat)[(((y-1)*ng2)+1):(((y-1)*ng2)+ng2)]
    mat[diff, group %in% paste0("group", 1:x)] <<- mat[diff, group %in% paste0("group", 1:x)] + rnorm(ng2, upreg, upregvar)
    return(diff)
  })
  
  mat[mat<0] <- 0
  mat <- round(mat)
  
  if(plot) {
    heatmap(mat, Rowv=NA, Colv=NA, col=colorRampPalette(c('blue', 'white', 'red'))(100), scale="none", ColSideColors=rainbow(G)[group], labCol=FALSE, labRow=FALSE)
  }
  
  return(list(mat=mat, group=group))
}

data <- simulate.data()
```

First, we will use a conventional PCA dimensionality approach followed by tSNE. What we get is a fuzzy blob. 

```{r}
myMudanObject <- Mudan$new("sim", data$mat, ncores=4)
myMudanObject$libSizeNormalize()
myMudanObject$varianceNormalize(plot=FALSE)
myMudanObject$dimensionalityReduction(nGenes = 1000, nPcs = 30, maxit=1000)
myMudanObject$getStandardEmbedding(plot=FALSE)

## PCA-based approach
plotEmbedding(myMudanObject$emb[['PCA']], groups=data$group, main="comb PCA")
```

Now we will use various graph-based community detection approaches to identify potential groups and the corresponding lower-dimensional space that can capture this group-separability. We will run tSNE on these lower-dimensional spaces corresponding to each set of identified groups.

```{r}
## community detection
myMudanObject$communityDetection(reductionType='pcs', communityName="Walktrap5", communityMethod=igraph::cluster_walktrap, k=5)
myMudanObject$communityDetection(reductionType='pcs', communityName="Walktrap10", communityMethod=igraph::cluster_walktrap, k=10)
myMudanObject$communityDetection(reductionType='pcs', communityName="Walktrap20", communityMethod=igraph::cluster_walktrap, k=20)

## use all communities detected
myMudanObject$modelCommunity()
myMudanObject$getMudanEmbedding(plot=FALSE)
plotEmbedding(myMudanObject$emb[['MUDAN']], groups=data$group, main="MUDAN")
```

We are able to much more clearly visually identify our original groups, even without prior knowledge of the groups. 

In theory, if we wanted to simply visualize known groups, we can also provide those group labels to identify a corresponding lower-dimensional space that can capture this group-separability. 

```{r}
myMudanObjectb <- Mudan$new("biased", data$mat, ncores=4)
myMudanObjectb$libSizeNormalize()
myMudanObjectb$varianceNormalize(plot=FALSE)
myMudanObjectb$dimensionalityReduction(nGenes = 1000, nPcs = 30, maxit=1000)
## use known groups rather than unbiased detection from graph-based approaches
myMudanObjectb$modelCommunity(groups=data$group)
myMudanObjectb$getMudanEmbedding(plot=FALSE)
plotEmbedding(myMudanObjectb$emb[['MUDAN']], groups=data$group, main="biased MUDAN")
```
